\documentclass[11pt]{article}


% Times font
%
\usepackage{newtxtext,newtxmath}
\usepackage[american]{babel}
\renewcommand\ttdefault{cmtt}


% Program environment
%
\usepackage{fancyvrb}
\newcommand{\progind}{\parindent}
\newcommand{\progformat}{\tt}
\newcommand{\progfontsize}{\small}
% \renewcommand{\progfontsize}{\footnotesize}
\DefineVerbatimEnvironment{program}{Verbatim}
   {xleftmargin=\progind,fontsize=\progfontsize,formatcom=\progformat,
    commandchars=\\\{\},samepage=true}
\DefineVerbatimEnvironment{programPageBreak}{Verbatim}
   {xleftmargin=\progind,fontsize=\progfontsize,formatcom=\progformat,
    commandchars=\\\{\}}
\DefineVerbatimEnvironment{programNoCmdChr}{Verbatim}
   {baselinestretch=1.0,xleftmargin=\progind,fontsize=\progfontsize,
    samepage=true}


% layout
%
\usepackage[top=1in, bottom=1in, left=1in, right=1in]{geometry}
\widowpenalty=0
\clubpenalty=0
\displaywidowpenalty=0
\raggedbottom
\sloppypar
\usepackage{needspace}
\def\denseitems{
  \itemsep1pt plus1pt minus1pt
  \parsep0pt plus0pt
  \parskip0pt\topsep0pt}



%%% Syntax macros
%%%
\newcommand{\prodName}[1]{&(\emph{#1})\\}
\newcommand{\nt}[1]{\textit{#1}}
\newcommand{\prog}[1]{\texttt{#1}}
\newcommand{\ORil}{\ \ |\ \ }
\newcommand{\IS}{&::=&}
\newcommand{\OR}{&|&}
\newcommand{\newsection}{\multicolumn{4}{l}{} \\}

\newcommand{\syntax}[2]{
  \needspace{3\baselineskip}
  \bigskip\par\noindent\textbf{#1}\par\smallskip%
  $\begin{array}{rcl@{\qquad}l}#2\end{array}$%
  \par\bigskip\noindent\ignorespaces
}

\newcommand{\lp}{\prog{(}}
\newcommand{\rp}{\prog{)}}
\newcommand{\paren}[1]{\lp#1\rp}
\newcommand{\pair}[2]{\paren{#1\prog{,}#2}}
\newcommand{\rbr}{\symbol{125}}
\newcommand{\bs}[1]{\symbol{92}}



\begin{document}

\title{BoGL Syntax}
\author{Martin Erwig}
\maketitle

\noindent
%
We generally adopt the Haskell rules for lexical syntax. In particular,
\nt{name} is used for function and parameter names and stands for alphanumeric
strings starting with a lowercase letter.
%
Similarly, \nt{Name} is used for symbols (which are basically nullary
constructors) and stands for alphanumeric strings starting with an uppercase
letter. It is also used for the names of games.
%
The nonterminal \nt{int} stands for integer constants.
%
Comments are also handled as in Haskell.

Unlike Haskell, we require all names (of values and types) to be defined before
they can be used in other definitions.

A game definition consists of a name, a number of type definitions, followed by
any number of value and function definitions. The type definitions must include
a definition for a board and an input type.
%
The board definition determines the board dimensions and the type of values
that can appear on a board. This type of board contents is given the type name
\prog{Piece}. Specifically, a user program that contains the  following type
definition:

\begin{program}
type Board = Array (n,m) of T
\end{program}
%
can be thought of as syntactic sugar for, and is thus equivalent to, the
following two definitions:

\begin{program}
type Piece = T
type Board = Array (n,m) of Piece
\end{program}
%
This definition allows us to assign the following type to the built-in function
place.

\begin{program}
place : (Piece,Board,Position) -> Board
\end{program}
%
% With this we can place integers and booleans and tuples, etc. on the board in a type-safe manner.

\syntax{Games and Type Definitions}{
\nt{game}    \IS  \prog{game}\ \nt{Name}
                  \ \nt{typedef}^*
                  \ \nt{board}\ \nt{input}
                  \ \nt{typedef}^*
                  \ \nt{valuedef}^*
                                                     \prodName{game definition}
% \nt{content} \IS  \prog{type Content = }\nt{type}      \prodName{input type}
\nt{board}   \IS  \prog{type Board = Array }\pair{\nt{int}}{\nt{int}}\
                  \prog{of}\ \nt{ptype}                \prodName{board type}
\nt{input}   \IS  \prog{type Input = }\nt{type}      \prodName{input type}
\nt{typedef} \IS  \prog{type }\nt{Name}\prog{ = }\nt{type}      \prodName{type definition}
}
%
Value definitions (for values and functions) require a signature declaring
their type plus an equation for defining the value. If a value is
parameterized, that is, if it is followed by a parameter list, it represents a
function.

A board definition (for a variable \nt{name} with declared type \prog{Board})
is equivalent to an array definition and may contain a number of definitions
for individual positions (given by pairs of integers) and for sets of
positions, which can be given by using a variable for either or both
coordinates. The semantics of a definition such as \prog{board!(x,2)~=~Empty}
is to set all fields in the second row to \prog{Empty}.

\syntax{Value Definitions}{
\nt{valuedef}  \IS \nt{signature}\ \nt{equation}   \prodName{value definition}
\nt{signature} \IS \nt{name}\ \prog{:}\ \nt{type}  \prodName{type signature}
\nt{equation}  \IS \nt{name}\ \prog{=}\ \nt{expr}  \prodName{value equation}
               \OR \nt{name}\ \nt{parlist}\ \prog{=}\ \nt{expr}
                   \prodName{function equation}
               \OR \nt{boarddef}^*                  \prodName{board definition}
\nt{boarddef}  \IS \nt{name}\,\prog{!}\,\pair{\nt{pos}}{\nt{pos}}\ \prog{=}\ \nt{expr}
                   \prodName{board equation}
\nt{pos}       \IS \nt{int} \ORil \nt{name}      \prodName{board positions}
\nt{parlist}   \IS \paren{\nt{par}_1\prog{,}\ \ldots\ \prog{,}\nt{par}_k}
                   \prodName{parameter list, $k\geq 1$}
\nt{par}       \IS \nt{name} \ORil \nt{parlist}   \prodName{parameter}
}
%
The type system builds on a collection of basic types, which include
predefined numbers and booleans as well as the type \prog{Symbol}, which is the
type for all symbol values (\nt{Name}).
%
% The predefined symbol values \prog{A} and \prog{B} also have the more specific type \prog{Player}.
%
The special type \prog{Input} is a synonym for the type of information that is
gathered from the user in each move during an execution of the game.
%
% Basic types also include types specific to the domain of board games. Among
% those, the type \prog{Position} is a synonym for \prog{(Int,Int)}. The type
% \prog{Positions} is an abstract data type for collections of positions, which
% can be though of as sets or lists of positions. However, the underlying
% representation is not exposed to the programmer. A value of type
% \prog{Positions} can only be generated and manipulated by a number of
% predefined operations.

An enumeration type is given by a set of symbols, and an  extended type adds to
a base type one or more symbols, which are values and can be thought of nullary
data constructors. For example, the Haskell type \prog{Maybe Int} can be
represented by the extended type \prog{Int~\&~\{Nothing\}}. Extended types
facilitate the extension of types by values for representing ``special''
situations. For example, \prog{Player~\&~\{Empty\}} is a type typically used in
a board type definition.

To avoid ambiguities and to simplify type checking we require that a symbol can
be used only in one enumeration type. The type of a symbol \prog{A} used in a definition \prog{type~T~=~U~\&~\{...,A,...\}} is \prog{T}.

A tuple type can contain extended types, and a plain type is either a tuple
type or an extended type. The language has first-order function types whose
argument and result types can be any plain type.

\syntax{Types}{
\nt{btype}    \IS  \prog{Bool} \ORil \prog{Int} \ORil \prog{Input}
                   % \prog{Symbol} \ORil \prog{Input}
                   % \ORil \prog{Void}
                                                         \prodName{base type}
              % \OR  \prog{Board} \ORil \prog{Player} \ORil
              %      \prog{Position} \ORil \prog{Positions}
              %                                              \prodName{game type}
\nt{etype}    \IS  \prog{\{}\nt{Name}_1\prog{,}\ \ldots\prog{,}\ \nt{Name}_k \prog{\}}               \prodName{enumeration type}
\nt{xtype}    \IS  \nt{btype} \ORil \nt{etype} \ORil
                   \nt{xtype}\ \prog{\&}\ \nt{etype}
                                                       \prodName{extended type}
\nt{ttype}    \IS  \paren{\nt{ptype}_1\prog{,}\ \ldots\prog{,}\ \nt{ptype}_k}
                                               \prodName{tuple type, $k\geq 2$}
\nt{ptype}     \IS  \nt{xtype} \ORil \nt{ttype}
                                                          \prodName{plain type}
\nt{ftype}    \IS  \nt{ptype}\ \prog{->}\ \nt{ptype}   \prodName{function type}
\nt{type}     \IS  \nt{ptype} \ORil \nt{ftype}                  \prodName{type}
}
%
Atomic expressions are either basic integer values or symbols. \prog{True} and
\prog{False} are symbols of type \prog{Bool}, all other symbols have either the
type \prog{Symbol} or the type of the type definition they occur in. For
example, the type definition \prog{type~Player~=~\{A,B\}} that occurs in the
BoGL prelude defines \prog{A} and \prog{B} to be symbols of type \prog{Player}.
%
Note that the case for infix application includes the Haskell notation for
array lookup \prog{board!(x,y)}.

\syntax{Expressions}{
\nt{expr}     \IS  \nt{int}                                 \prodName{integer}
              \OR  \nt{Name}                                 \prodName{symbol}
              \OR  \nt{name}                               \prodName{variable}
            % \OR  \prog{A} \ORil \prog{B}                   \prodName{player}
              \OR  \paren{\nt{expr}}       \prodName{parenthesized expression}
              \OR  \paren{\nt{expr}_1\prog{,}\ \ldots\ \prog{,}\nt{expr}_k}
                                                    \prodName{tuple, $k\geq 2$}
%%% function application
              \OR  \nt{name}\paren{\nt{expr}_1\prog{,}\ \ldots\
                                              \prog{,}\nt{expr}_k}
                                               \prodName{function application}
              \OR  \nt{expr}\ \nt{binop}\ \nt{expr}
                                                  \prodName{infix application}
%%% bindings
             \OR  \prog{let}\ \nt{name}\ \prog{=}\ \nt{expr}\
                   \prog{in}\ \nt{expr}            \prodName{local definition}
%%% control strructures
              \OR  \prog{if}\ \nt{expr}\ \prog{then}\ \nt{expr}\
                   \prog{else}\ \nt{expr}               \prodName{conditional}
              \OR  \prog{while}\ \nt{expr}\ \prog{do}\ \nt{expr}
                                                         \prodName{while loop}
%
\nt{binop}    \IS  \prog{+} \ORil \prog{-} \ORil
                   \prog{==} \ORil \prog{!} \ORil \ldots
                                                   \prodName{binary operation}
}
%
Here is a list of \emph{required built-in} types and functions.
%
Note the following.
%
\begin{itemize}\denseitems
%
\item The type definition for \prog{Position} is for convenience; it allows us
to give the shown type for the function \prog{place}. An implementation could
omit this definition and instead use \prog{(Int,Int)}.
%
\item Despite its type, \prog{input} is a function that triggers user-input
requests and returns the values given by the user, which have to be of type
\prog{Input}, which is a user defined type.
%
\item The first argument type of the function \prog{place} is \prog{Symbol},
which is imprecise (it's a sloppy approach to model a form of ad hoc
polymorphism). The type checker should make sure that \prog{place} receives as
first arguments only values of the same type that has been used in the type
definition for \prog{Board}.
%
The same remark applies to the \prog{Symbol} type argument of the functions
\prog{inARow}, \prog{countBoard}, \prog{countRow}, and \prog{countColumn}.
%
\end{itemize}

\begin{program}
type Position = (Int,Int)

input       : Input
place       : (Piece,Board,Position) -> Board
inARow      : (Int,Symbol,Board) -> Bool
countBoard  : (Symbol,Board) -> Int
countRow    : (Symbol,Board,Int) -> Int
countColumn : (Symbol,Board,Int) -> Int
\end{program}
%
Here is the default content of the file \prog{Prelude.bgl}.

\begin{programPageBreak}
-- Game types
--
type Player = {A,B}
type State = (Board,Player)


-- Players
--
goFirst : Player
goFirst = A

next : Player -> Player
next(p) = if p == A then B else A


-- Initial board and board operations
--
initialBoard : Board
initialBoard(x,y) = Empty

isValid : (Board,Position) -> Bool
isValid(b,p) = if b!p == Empty then True else False

isFull : Board -> Bool
isFull(b) = countBoard(Empty,b) == 0


-- Game state operations
--
tryMove : State -> State
tryMove(b,p) = let pos = input in
                   if isValid(b,pos) then (place(p,b,pos),next(p))
                                     else (b,p)
\end{programPageBreak}

\end{document}


\begin{programPageBreak}
--
-- BoGL general definitions
--


-- Game types
--
type Player = {A,B}
type State = (Board,Player)


-- Players
--
goFirst : Player
goFirst = A

next : Player -> Player
next(A) = B
next(B) = A


-- Initial board and board operations
--
initialBoard : Board
initialBoard(x,y) = Empty

isValid : (Board,Position) -> Bool
isValid(b,p) = if b!p == Empty then True else False

isFull : Board -> Bool
isFull(b) = countBoard(Empty,b) == 0


-- Game state operations
--
tryMove : State -> State
tryMove(b,p) = let pos = input in
                   if isValid(b,pos) then (place(p,b,pos),next(p))
                                     else (b,p)
\end{programPageBreak}
